\chapter{Use Case: EPOS API}
\label{chap:caso-epos}

\section{Introduction}
\label{chap5:sec:intro}

This chapter describes the application of the framework introduced
in Chapter~\ref{chap:metodo-arq} to the production \ac{API} of the
laboratory, \texttt{api\_v4.0}. Section~\ref{sec:epos-context}
explains briefly the \ac{EPOS} context that the \ac{API} serves.
Section~\ref{sec:apiv4-arch} walks through the structure of the
application itself, with emphasis on the elements that the testing
framework has to be aware of (resource classes, output formats,
persistence). Section~\ref{sec:pipeline} dissects the
\texttt{.gitlab-ci.yml} file, job by job. Sections~\ref{sec:diffharness}
and~\ref{sec:httpdiff} document the two diff harnesses, code-level
and \ac{HTTP}-level, including their inputs, outputs and reporting
contract. Section~\ref{sec:reports} discusses how the reports are
consumed and the kinds of issues the harnesses have surfaced in
practice. Section~\ref{chap5:sec:concs} concludes.

\section{The \ac{EPOS} Context}
\label{sec:epos-context}

The \ac{EPOS}~\cite{epos} is a pan-European research infrastructure
in solid Earth science, organised into Thematic Core Services for
seismology, near-fault observatories, volcano observations,
geomagnetic observations, anthropogenic hazards and \ac{GNSS} data
and products~\cite{eposGnss}, among others. The \ac{SEGAL}
laboratory of \ac{UBI} contributes to the \ac{GNSS} Thematic Core
Service by operating a service node that hosts station metadata and
\ac{RINEX}~\cite{rinex} observation files. The production \ac{API}
described in this chapter is the public interface of that service
node.

From the point of view of the test framework, the salient properties
of this domain are three. First, the entities are relatively static
(new stations are added rarely, file metadata grows over time but
in well-defined ways), which makes a snapshot of the database a
stable baseline against which to compare. Second, the typical client
request returns a long list of records, which means that any
behavioural divergence produces a long diff and the comparison
algorithm must avoid being overwhelmed by it. Third, the same
information is offered in multiple serialization formats, which
multiplies the surface area to be tested but does not change the
underlying logic.

\section{Architecture of \texttt{api\_v4.0}}
\label{sec:apiv4-arch}

\subsection{Source Layout}

The application is a Maven \ac{WAR} project targeting Java SE 21 and
Jakarta EE 11, deployed onto Payara Server 6 with
\texttt{finalName=ROOT} (so that it is reachable at the root context
\texttt{/}). Its source tree is organised into four top-level
packages, summarised in Table~\ref{tab:apiv4-packages}.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{l p{8.5cm}}
    \toprule
    \textbf{Package} & \textbf{Purpose} \\
    \midrule
    \texttt{Config}    & Database connection bootstrap
                         (\texttt{Loader.java}, persistence-unit
                         factory). \\
    \texttt{DB\_Tables} & The forty \ac{JPA} entity classes that
                         model the relational schema (stations,
                         networks, agencies, file types, observation
                         summaries, embargos, \ldots). \\
    \texttt{Geometry}  & Geometric validation, in particular the
                         native polygon checks used by the
                         bounding-box queries. \\
    \texttt{Glass}     & The \ac{REST} layer proper: \ac{JAX-RS}
                         resources (\texttt{StationEndpoints},
                         \texttt{FilesEndpoints}, \texttt{ListEndpoints},
                         \texttt{Start}), parsers, format writers
                         (\ac{JSON}/\ac{XML}/\ac{CSV}/script) and
                         the validation/shape helpers
                         (\texttt{Validate.java},
                         \texttt{Shape.java}). \\
    \bottomrule
  \end{tabular}
  \caption{Top-level packages of the production \ac{API}.}
  \label{tab:apiv4-packages}
\end{table}

The four packages cooperate as a classical layered architecture
(Figure~\ref{fig:apiv4-layers}): the \texttt{Glass} layer exposes the
\ac{HTTP} surface, the validation and geometry helpers form a thin
business layer, the \ac{JPA} entities of \texttt{DB\_Tables} act as
the persistence layer, and \texttt{Config} is a cross-cutting concern
that handles bootstrap.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\small},
      node distance=4mm,
      layer/.style={draw, rounded corners=2pt, align=center, thick,
                    minimum width=92mm, minimum height=12mm},
      cc/.style={draw, dashed, rounded corners=2pt, align=center,
                 minimum width=28mm, minimum height=10mm,
                 font=\footnotesize},
      arrow/.style={-{Stealth[length=2mm]}, thick}
    ]
    \node[layer, fill=dkblue!10] (rest)
      {\textbf{REST layer}, package \texttt{Glass} \\
       \footnotesize Start, StationEndpoints, FilesEndpoints, ListEndpoints, Format writers};
    \node[layer, fill=mauve!10, below=of rest] (biz)
      {\textbf{Business / validation layer} \\
       \footnotesize \texttt{Validate.java}, \texttt{Shape.java}, package \texttt{Geometry} (\texttt{NativePolygonCheck})};
    \node[layer, fill=dkgreen!10, below=of biz] (jpa)
      {\textbf{Persistence layer}, package \texttt{DB\_Tables} \\
       \footnotesize 40 JPA entities (Station, RinexFile, Network, Agency, \ldots)};
    \node[layer, fill=dark!10, below=of jpa] (db)
      {\textbf{Database}, PostgreSQL \texttt{192.168.168.30:5432/DB\_3.0\_J} \\
       \footnotesize over the SEGAL internal network / \ac{VPN}};

    \node[cc, right=6mm of biz, yshift=8mm] (cfg)
      {\textbf{Config} \\\texttt{Loader.java} \\ env / file / default};
    \node[cc, right=6mm of biz, yshift=-8mm] (pu)
      {Persistence unit \\\texttt{MeuPU} \\ HikariCP pool};

    \draw[arrow] (rest) -- (biz);
    \draw[arrow] (biz)  -- (jpa);
    \draw[arrow] (jpa)  -- (db);
    \draw[arrow, dashed] (cfg) -- (db);
    \draw[arrow, dashed] (pu)  -- (jpa);
  \end{tikzpicture}%
  }
  \caption{Layered architecture of \texttt{api\_v4.0}. Solid arrows
  denote ordinary call/data flow; dashed arrows denote configuration
  and infrastructure binding established at deployment time.}
  \label{fig:apiv4-layers}
\end{figure}

\subsection{Endpoints and Output Formats}

The \ac{REST} surface is declared in four classes inside the
\texttt{Glass} package. \texttt{Start.java} carries the
\texttt{@ApplicationPath("/api")} annotation and registers a
\texttt{PingResource} for the trivial \texttt{/api/ping} liveness
check used by both the diff harness and the deploy verification of
\texttt{http-diff-job}. The three substantive resources are listed
in Table~\ref{tab:endpoints}.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{l l p{4.5cm}}
    \toprule
    \textbf{Resource} & \textbf{Base path} & \textbf{Approximate scope} \\
    \midrule
    \texttt{StationEndpoints} & \texttt{/api/stations}
       & Seven endpoints for station metadata, indexed by station
         identifier, location, agency, network, equipment and
         coordinates. \\
    \texttt{FilesEndpoints}   & \texttt{/api/files}
       & Four endpoints for \ac{RINEX} file metadata, by agency,
         antenna, bounding box and date; by marker; high-rate;
         and an aggregated combination. \\
    \texttt{ListEndpoints}    & \texttt{/api/list}
       & A parametric endpoint that returns the valid values of a
         given filter (agencies, networks, countries, equipment
         types, file types). \\
    \bottomrule
  \end{tabular}
  \caption{The three substantive \ac{JAX-RS} resources of
  \texttt{api\_v4.0}.}
  \label{tab:endpoints}
\end{table}

Every endpoint accepts a \texttt{\{format\}} path parameter and
emits the response in \ac{JSON}, \ac{XML}, \ac{CSV} or, for the
file endpoints, a downloadable shell script. The same \ac{OpenAPI}
3.1.0 document is served at \texttt{/swagger.html} through a Swagger
\ac{UI} hosted inside the \ac{WAR}.

Figure~\ref{fig:reqlifecycle} traces the end-to-end path of a single
request through the Payara process, from the \ac{HTTP} listener down
to the database and back, calling out the Jakarta specifications
involved at each step.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\footnotesize},
      node distance=4mm,
      flowstep/.style={draw, rounded corners=2pt, align=center, thick,
                   minimum width=70mm, minimum height=9mm},
      spec/.style={font=\scriptsize\itshape, text=dkblue},
      arrow/.style={-{Stealth[length=2mm]}, thick}
    ]
    \node[flowstep, fill=lgray]      (client)  {\textbf{HTTP client} \\ \texttt{GET /api/stations/agency/IGS/json}};
    \node[flowstep, fill=dkblue!10, below=of client]   (grizzly) {\textbf{Grizzly HTTP listener (Payara)}};
    \node[flowstep, fill=dkblue!10, below=of grizzly]  (servlet) {\textbf{Servlet container}};
    \node[flowstep, fill=mauve!10, below=of servlet]   (dispatch){\textbf{JAX-RS dispatcher}, matches \texttt{@Path} and verb};
    \node[flowstep, fill=mauve!10, below=of dispatch]  (cdi)     {\textbf{Resource bean}\ \texttt{StationEndpoints} \\ \texttt{@Inject} resolved by CDI container};
    \node[flowstep, fill=mauve!10, below=of cdi]       (valid)   {\textbf{Validate.java}, parses and checks query parameters};
    \node[flowstep, fill=dkgreen!10, below=of valid]   (em)      {\textbf{EntityManager} (Hibernate, persistence unit \texttt{MeuPU})};
    \node[flowstep, fill=dkgreen!10, below=of em]      (jdbc)    {\textbf{HikariCP} pool \(\rightarrow\) JDBC};
    \node[flowstep, fill=dark!10,   below=of jdbc]     (db)      {\textbf{PostgreSQL} returns rows};
    \node[flowstep, fill=dkgreen!10, below=of db]      (pojo)    {Rows hydrated into \textbf{JPA entities}};
    \node[flowstep, fill=mauve!10, below=of pojo]      (writer)  {\textbf{Format writer} chosen by \texttt{\{format\}} (JSON / XML / CSV / script)};
    \node[flowstep, fill=dkblue!10, below=of writer]   (resp)    {Response serialized and returned to the listener};
    \node[flowstep, fill=lgray, below=of resp]         (cli2)    {\textbf{HTTP client} receives \texttt{200 OK} body};

    \draw[arrow] (client)   -- (grizzly);
    \draw[arrow] (grizzly)  -- (servlet);
    \draw[arrow] (servlet)  -- (dispatch);
    \draw[arrow] (dispatch) -- (cdi);
    \draw[arrow] (cdi)      -- (valid);
    \draw[arrow] (valid)    -- (em);
    \draw[arrow] (em)       -- (jdbc);
    \draw[arrow] (jdbc)     -- (db);
    \draw[arrow] (db)       -- (pojo);
    \draw[arrow] (pojo)     -- (writer);
    \draw[arrow] (writer)   -- (resp);
    \draw[arrow] (resp)     -- (cli2);

    \node[spec, right=2mm of grizzly]  {Jakarta Servlet};
    \node[spec, right=2mm of dispatch] {Jakarta REST (JAX-RS)};
    \node[spec, right=2mm of cdi]      {Jakarta CDI};
    \node[spec, right=2mm of em]       {Jakarta Persistence (JPA)};
    \node[spec, right=2mm of writer]   {Jackson / JSON-B};
  \end{tikzpicture}%
  }
  \caption{End-to-end request lifecycle of a \ac{JAX-RS} call to
  \texttt{api\_v4.0} on Payara. The right column names the Jakarta
  specification active at each step.}
  \label{fig:reqlifecycle}
\end{figure}

\subsection{Persistence}

The persistence unit, declared in
\texttt{src/main/resources/META-INF/persistence.xml}, is named
\texttt{MeuPU} (inherited from the legacy code). It is configured as
\texttt{RESOURCE\_LOCAL}, with Hibernate as the \ac{JPA} provider
and a connection pool managed by HikariCP\@. Database connection
parameters, host, port, database name, user, password, are
resolved at startup by \texttt{Config/Loader.java}, with the
following precedence: environment variables, then a
\texttt{Config.config} file at the root of the deployment, then
hard-coded defaults pointing at the laboratory's database
(\texttt{192.168.168.30:5432}).

The laboratory maintains two database fixtures with the same schema
but different volumes of data:

\begin{itemize}
  \item \texttt{DB\_3.0\_J} hosts a small but representative production
        subset, on the order of ten stations, eighty-six networks,
        two hundred and thirty agencies and roughly one hundred and
        forty-six thousand \texttt{rinex\_file} rows at the time of
        writing. This is the fixture used by the correctness diff
        runs, because it is small enough to allow per-case comparison
        of full response bodies without overwhelming the report.
  \item \texttt{DB\_3.0\_F} hosts the full dataset and is used by the
        volume-oriented diff runs, where the goal is not byte-for-byte
        equality but the detection of changes that would only be
        apparent at scale, for example, an inadvertent
        \texttt{N+1} query introduced by a refactor, or a query plan
        regression visible only on a large table.
\end{itemize}

The precedence rules of \texttt{Loader.java} are what allow the
\texttt{http-diff-job} to switch the running \acp{WAR} between the two
fixtures without rebuilding the artifact: the job sets the
appropriate environment variables in its \texttt{variables:} block
and the application picks them up at boot.

The full schema, as introspected from the running database, contains
forty-two tables in the \texttt{public} schema and around fifty
foreign-key constraints, organised into a few clear groups: core
domain (\texttt{station}, \texttt{network}, \texttt{agency},
\texttt{country}, \texttt{rinex\_file}, \texttt{highrate\_rinex\_file},
\texttt{file\_type}), junction tables (\texttt{agency\_network},
\texttt{station\_network}, \texttt{station\_agency\_role},
\texttt{node\_station}, \texttt{node\_data\_center}), quality-control
summaries (\texttt{qc\_report\_summary},
\texttt{qc\_constellation\_summary},
\texttt{qc\_header\_observations}, \texttt{qc\_navigation\_msg} and
four \texttt{qc\_observation\_summary\_\{c,d,l,s\}} variants), and
auxiliary tables for embargoes, documents, logs, licenses and
infrastructure. A simplified view of the central entities is given
in Figure~\ref{fig:datamodel}; the chapter on technologies
(Section~\ref{sec:epos-db}) discusses the design of the database
itself in more detail.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\footnotesize},
      entity/.style={draw, rounded corners=2pt, align=center, thick,
                     minimum width=24mm, minimum height=11mm,
                     fill=dkgreen!10},
      lookup/.style={entity, fill=mauve!10},
      junc/.style={entity, fill=lgray, minimum width=22mm},
      arrow/.style={-{Stealth[length=2mm]}, thick},
      rel/.style={font=\scriptsize, text=dark}
    ]
    \node[entity] (station)
      {\textbf{station} \\\scriptsize id, marker[9], lat, lon, x/y/z,\\\scriptsize country\_code, station\_type};
    \node[junc, left=10mm of station]   (sn) {\textbf{station\_network} \\\scriptsize include\_date};
    \node[entity, left=10mm of sn]      (network) {\textbf{network} \\\scriptsize id, name, network\_type, doi};
    \node[junc, left=10mm of network]   (an) {\textbf{agency\_network} \\\scriptsize role};
    \node[lookup, left=10mm of an]      (agency)  {\textbf{agency} \\\scriptsize id, name, abbreviation,\\\scriptsize ror};
    \node[lookup, above=8mm of station] (country) {\textbf{country} \\\scriptsize iso\_code (PK), name};
    \node[entity, right=12mm of station](rinex)   {\textbf{rinex\_file} \\\scriptsize id, name, file\_size,\\\scriptsize reference\_date, md5checksum};
    \node[entity, below=8mm of rinex]   (highrate){\textbf{highrate\_rinex\_file}};
    \node[lookup, right=12mm of rinex]  (ftype)   {\textbf{file\_type} \\\scriptsize format, sampling\_window};
    \node[junc, below=8mm of station]   (nodest)  {\textbf{node\_station}};
    \node[entity, below=8mm of nodest]  (node)    {\textbf{node} / \textbf{data\_center}};

    \draw[arrow] (agency)  -- node[rel,above]{1..*} (an);
    \draw[arrow] (network) -- node[rel,above]{1..*} (an);
    \draw[arrow] (station) -- node[rel,above]{1..*} (sn);
    \draw[arrow] (network) -- node[rel,above]{1..*} (sn);
    \draw[arrow] (country) -- node[rel,right]{1..*} (station);
    \draw[arrow] (station) -- node[rel,above]{1..*} (rinex);
    \draw[arrow] (station) -- node[rel,above,sloped]{1..*} (highrate);
    \draw[arrow] (ftype)   -- node[rel,above]{1..*} (rinex);
    \draw[arrow] (ftype)   -- node[rel,below,sloped]{1..*} (highrate);
    \draw[arrow] (station) -- node[rel,right]{1..*} (nodest);
    \draw[arrow] (nodest)  -- node[rel,right]{*..1} (node);
  \end{tikzpicture}%
  }
  \caption{Simplified view of the core \ac{EPOS} data model, with
  representative columns as they appear in the live PostgreSQL schema.
  Green entities carry domain data; purple entities are
  controlled-vocabulary lookups (note that \texttt{country.iso\_code}
  is the natural primary key, not a surrogate \texttt{id});
  grey entities are junction tables that materialise the many-to-many
  relationships of the model. Arrows point from the ``one'' side to
  the ``many'' side of the relationship.}
  \label{fig:datamodel}
\end{figure}

\section{The \ac{CI}/\ac{CD} Pipeline}
\label{sec:pipeline}

The pipeline of \texttt{api\_v4.0} extends the patterns validated on
the first prototype with two additional jobs in a new \texttt{diff}
stage. Each diff job compares the current build against \emph{two}
baselines, in what amounts to a three-way comparison: the immediate
parent commit \texttt{HEAD\textasciitilde 1} (which catches
regressions introduced by the commit currently being pushed) and a
\emph{fixed} reference commit identified by a SHA stored in the
\texttt{DIFF\_FIXED\_BASELINE} GitLab \ac{CI}/\ac{CD} variable
(which catches the slow drift that accumulates over many commits
along a long-running refactor branch). The full file is shown in
Listing~\ref{lst:apiv4-ci-1} and Listing~\ref{lst:apiv4-ci-2}.

\begin{lstlisting}[language=yaml,caption={Header, build and integration test jobs of the production pipeline.},label=lst:apiv4-ci-1]
default:
  image: maven:3.9.6-eclipse-temurin-21
  tags:
    - ubi-runner-.31

cache:
  key: global-maven-cache
  paths:
    - .m2/repository

variables:
  MAVEN_OPTS: "-Dmaven.repo.local=$CI_PROJECT_DIR/.m2/repository"
  GIT_DEPTH: "20"

stages:
  - build
  - test
  - diff

build-job:
  stage: build
  script:
    - mvn clean package -DskipTests
  artifacts:
    paths:
      - target/*.war

integration-test-job:
  stage: test
  script:
    - mvn verify
  artifacts:
    when: always
    reports:
      junit:
        - "target/failsafe-reports/failsafe-summary.xml"
        - "target/failsafe-reports/TEST-*.xml"
    paths:
      - target/failsafe-reports/
    expire_in: 1 week
\end{lstlisting}

The pinned runner tag, the cached Maven repository and the
\texttt{maven:\allowbreak 3.9.6-eclipse-\allowbreak temurin-21} base image are direct
descendants of the first prototype. Two additions are worth
calling out. The \texttt{GIT\_DEPTH: "20"} variable bumps the
default shallow-clone depth from one to twenty commits, which is
the minimum required for the diff jobs to be able to inspect
\texttt{HEAD\textasciitilde 1} without an additional
\texttt{git fetch}. The third stage, \texttt{diff}, hosts the two
new jobs.

\begin{lstlisting}[language=yaml,caption={\texttt{validate-diff-job} and \texttt{http-diff-job}, the three-way differential testing jobs.},label=lst:apiv4-ci-2]
validate-diff-job:
  stage: diff
  needs:
    - job: build-job
      artifacts: false
  variables:
    DIFF_FIXED_BASELINE: ""   # commit SHA, set in GitLab CI/CD Variables
  script:
    - |
      run_against_baseline() {
        local LABEL="$1"; local BASELINE_SHA="$2"
        if [ -z "$BASELINE_SHA" ]; then
          echo "No $LABEL baseline available - skipping"
          return 0
        fi
        local WORK="/tmp/${LABEL}-baseline-build"
        git worktree remove --force "$WORK" 2>/dev/null || true
        git worktree prune
        rm -rf "$WORK"
        git worktree add "$WORK" "$BASELINE_SHA"
        ( cd "$WORK" && mvn -q -o package -DskipTests )
        export BASELINE_WAR="$WORK/target/ROOT.war"
        export DIFF_REPORT_DIR="target/diff-report/$LABEL"
        export DIFF_LABEL="$LABEL"
        mkdir -p "$DIFF_REPORT_DIR"
        mvn test -Dtest=DiffHarness -DfailIfNoTests=false
      }
      RECENT_SHA=$(git rev-parse --verify HEAD~1 2>/dev/null || echo "")
      run_against_baseline recent "$RECENT_SHA"
      run_against_baseline fixed  "$DIFF_FIXED_BASELINE"
  artifacts:
    when: always
    paths:
      - target/diff-report/
    reports:
      junit:
        - "target/surefire-reports/TEST-Diff.DiffHarness.xml"
    expire_in: 4 weeks

http-diff-job:
  stage: diff
  needs:
    - job: build-job
      artifacts: true
  variables:
    DATABASE_URL: "postgres://postgres:postgres@192.168.168.30:5432/DB_3.0_J"
    JAKARTA_SCHEMA_ACTION: "none"
    DIFF_FIXED_BASELINE: ""
  script:
    - |
      build_baseline() {
        local LABEL="$1"; local BASELINE_SHA="$2"
        local WORK="/tmp/${LABEL}-http-baseline-build"
        git worktree remove --force "$WORK" 2>/dev/null || true
        git worktree prune
        rm -rf "$WORK"
        git worktree add "$WORK" "$BASELINE_SHA"
        ( cd "$WORK" && mvn -q -o package -DskipTests )
        echo "$WORK/target/ROOT.war"
      }
      RECENT_SHA=$(git rev-parse --verify HEAD~1 2>/dev/null || echo "")
      [ -n "$RECENT_SHA" ]            && export BASELINE_RECENT_WAR=$(build_baseline recent "$RECENT_SHA")
      [ -n "$DIFF_FIXED_BASELINE" ]   && export BASELINE_FIXED_WAR=$(build_baseline fixed  "$DIFF_FIXED_BASELINE")
      mvn test -Dtest=HttpDiff -DfailIfNoTests=false
  artifacts:
    when: always
    paths:
      - target/http-diff-report/
    reports:
      junit:
        - "target/surefire-reports/TEST-Diff.HttpDiff.xml"
    expire_in: 4 weeks
\end{lstlisting}

A few design decisions visible in the two listings deserve a comment.

\paragraph{Two baselines, one job.}
Each diff job runs the harness twice, once per baseline, with the
report written into a separate subdirectory of
\texttt{target/diff-report/} (or \texttt{target/http-diff-report/}).
The \emph{recent} baseline is the immediate parent of \texttt{HEAD},
which keeps the per-commit comparison focused on the change just
introduced and avoids the avalanche of diffs that would otherwise
occur on long-running refactor branches. The \emph{fixed} baseline
is a commit \ac{ID} stored in the GitLab variable
\texttt{DIFF\_FIXED\_BASELINE}: it acts as a long-term reference
against which the cumulative drift of the branch is measured, and is
typically pointed at the last commit known to have shipped to the
\ac{EPOS} portal. The two baselines surface different classes of
regressions (one immediate, one accumulated) without doubling the
maintenance cost of test fixtures.

\paragraph{Worktree-based baseline build.}
The baseline \ac{WAR} is built by checking out the baseline commit
into a separate Git worktree (\texttt{/tmp/baseline-build}) and
invoking \texttt{mvn -q -o package -DskipTests} there. Worktrees
were chosen over the more obvious \texttt{git checkout} because they
preserve the current working directory untouched: the harness can
freely compare files from both checkouts without having to alternate
between branches.

\paragraph{Defensive skip per baseline.}
Each baseline is treated independently. If \texttt{HEAD\textasciitilde 1}
does not exist (the trivial case of the first commit of the
repository), the \emph{recent} comparison is skipped. If
\texttt{DIFF\_FIXED\_BASELINE} is unset, which is the expected
state at the very start of a refactor branch, before a fixed
reference has been chosen, the \emph{fixed} comparison is
skipped. A pipeline can therefore have zero, one or two diff sub-runs
depending on the situation, and never fails simply because a
baseline is unavailable.

\paragraph{Job dependencies.}
\texttt{validate-diff-job} does not need the artifact of
\texttt{build-job}, it rebuilds the current \ac{WAR} as a side
effect of \texttt{mvn test}, and is therefore declared with
\texttt{needs: artifacts: false} to avoid an unnecessary artifact
download. \texttt{http-diff-job}, on the other hand, requires the
current \ac{WAR} to spawn Payara Micro and so declares
\texttt{artifacts: true}.

\section{The Code-Level Diff: \texttt{DiffHarness.java}}
\label{sec:diffharness}

\texttt{DiffHarness.java} is a JUnit~5 test that compares two
versions of a specific Java class on a shared input grid, in
process. Its core idea is to load the baseline and the \ac{HEAD}
\acp{WAR} through two independent \texttt{URLClassLoader} instances,
to reflect into the class under test on both sides and to invoke
each method of interest with the same arguments. Any difference
between the two returned values, thrown exceptions included,
is recorded as a row in the diff report.

The pipeline invokes the harness once per baseline (recent and
fixed; see Section~\ref{sec:pipeline}). The exported environment
variables \texttt{BASELINE\_WAR}, \texttt{DIFF\_REPORT\_DIR} and
\texttt{DIFF\_LABEL} tell the harness which \ac{WAR} to load against
\ac{HEAD} and where to write its outputs. The on-disk layout is
\texttt{target/diff-report/recent/} and
\texttt{target/diff-report/fixed/}, so a glance at the artifact tree
of a finished pipeline immediately distinguishes a per-commit
regression from a long-term drift.

The harness focuses on two classes for which behaviour is critical
and for which an entire endpoint cannot be exercised in isolation:
\texttt{Validate}, which parses and validates client query
parameters, and \texttt{Shape}, which evaluates geometric polygon
queries against the bounding box of a request. Both classes are
purely functional, take simple inputs (strings, lists of doubles)
and return simple outputs (booleans, lists, error strings), which
makes them well suited to a code-level diff.

The output of each sub-run is twofold. A \ac{JUnit} XML report
(\texttt{TEST-Diff.\allowbreak DiffHarness.xml}, with the \texttt{DIFF\_LABEL}
appearing as a test-suite property) integrates with GitLab so that
divergences appear as failed assertions on the pipeline page. In
parallel, a richer \ac{CSV}/\ac{JSON} report under the per-baseline
subdirectory records every input row, the value returned by the
baseline, the value returned by \ac{HEAD} and an indicator of
whether the two match. The \ac{CSV} format makes the report easy
to open in a spreadsheet for ad-hoc analysis.

\section{The \ac{HTTP}-Level Diff: \texttt{HttpDiff.java}}
\label{sec:httpdiff}

\texttt{HttpDiff.java} is the second harness, and the one that
exercises the \ac{API} end-to-end. Its execution proceeds in four
phases, summarised graphically in Figure~\ref{fig:httpdiff-phases}.
Unlike the code-level harness, which is invoked twice by the
pipeline, the \ac{HTTP} harness runs once and internally boots
\emph{all} the available \acp{WAR} in parallel (the \ac{HEAD}
\ac{WAR}, the recent baseline if any, and the fixed baseline if
any). It then issues every test case against every running instance
and emits one diff report per baseline.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\footnotesize},
      proc/.style={draw, rounded corners=2pt, align=center, thick,
                   minimum width=26mm, minimum height=10mm,
                   fill=dkblue!10},
      micro/.style={proc, fill=mauve!10, minimum width=30mm},
      head/.style={proc, fill=dkgreen!10, minimum width=30mm},
      stage/.style={font=\small\bfseries, text=dark},
      barrow/.style={-{Stealth[length=2mm]}, thick},
      darrow/.style={-{Stealth[length=2mm]}, thick, dashed}
    ]
    \node[proc] (harness) {\textbf{HttpDiff.java} \\ \footnotesize (JUnit 5)};
    \node[head,  right=18mm of harness] (p80) {\textbf{payara-micro :8080} \\ \footnotesize HEAD WAR};
    \node[micro, right=8mm  of p80]     (p81) {\textbf{payara-micro :8081} \\ \footnotesize recent baseline};
    \node[micro, right=8mm  of p81]     (p82) {\textbf{payara-micro :8082} \\ \footnotesize fixed baseline};
    \node[proc, fill=dark!10, right=12mm of p82] (pg) {\textbf{PostgreSQL} \\ \footnotesize 192.168.168.30};

    \node[stage, below=4mm of harness] (s1) {Phase 1, Boot};
    \draw[barrow] ($(harness.south)+(0,-12mm)$) -| (p80.south);
    \draw[barrow] ($(harness.south)+(0,-12mm)$) -| (p81.south);
    \draw[barrow] ($(harness.south)+(0,-12mm)$) -| (p82.south);

    \node[stage, below=20mm of s1] (s2) {Phase 2, Verify};
    \draw[barrow] ($(harness.south)+(0,-30mm)$) -- ++(1.8,0)
        node[midway,above,font=\scriptsize]{GET /api/ping} -| (p80);
    \draw[barrow] ($(harness.south)+(0,-34mm)$) -- ++(4.6,0)
        node[midway,above,font=\scriptsize]{GET /api/ping} -| (p81);
    \draw[barrow] ($(harness.south)+(0,-38mm)$) -- ++(7.4,0)
        node[midway,above,font=\scriptsize]{GET /api/ping} -| (p82);

    \node[stage, below=24mm of s2] (s3) {Phase 3, Replay \& 3-way diff};
    \draw[barrow] ($(harness.south)+(0,-56mm)$) -- ++(1.8,0)
        node[midway,above,font=\scriptsize]{case[i]} -| (p80);
    \draw[barrow] ($(harness.south)+(0,-60mm)$) -- ++(4.6,0)
        node[midway,above,font=\scriptsize]{case[i]} -| (p81);
    \draw[barrow] ($(harness.south)+(0,-64mm)$) -- ++(7.4,0)
        node[midway,above,font=\scriptsize]{case[i]} -| (p82);
    \draw[darrow] ($(p80.south)+(0,-12mm)$) -- ($(pg.south)+(0,-12mm)$)
        node[midway,below,font=\scriptsize]{all three query the same DB};

    \node[proc, fill=dkgreen!10, below=70mm of harness, minimum width=132mm]
        (diff) {\textbf{Diff} HEAD vs recent \(\rightarrow\) \texttt{http-diff-report/recent/} \quad
                \textbf{Diff} HEAD vs fixed \(\rightarrow\) \texttt{http-diff-report/fixed/}};

    \node[stage, below=4mm of diff] (s4) {Phase 4, Teardown + Report};
    \draw[barrow] ($(harness.south)+(0,-92mm)$) -- ++(10.0,0)
        node[midway,above,font=\scriptsize]{SIGTERM all payara-micro};
  \end{tikzpicture}%
  }
  \caption{Phases of \texttt{HttpDiff.java}. The harness boots one
  Payara Micro per available \ac{WAR}, the \ac{HEAD} build and
  whichever of the recent/fixed baselines were provided, waits
  for all of them to be ready, fires every test case against every
  instance in parallel, then performs a three-way comparison and
  emits one diff report per baseline. All instances share the same
  real PostgreSQL fixture over the laboratory \ac{VPN}.}
  \label{fig:httpdiff-phases}
\end{figure}

\paragraph{Phase 1, Boot.}
The harness spawns up to three child \ac{JVM} processes from the
Payara Micro \ac{JAR} that the Maven build had previously downloaded
into \texttt{target/payara-micro/}. The current \ac{WAR}
(\texttt{target/ROOT.war}) is deployed on port 8080. If the
\texttt{BASELINE\_RECENT\_WAR} environment variable is set, the
recent baseline is deployed on port 8081; if
\texttt{BASELINE\_FIXED\_WAR} is set, the fixed baseline is deployed
on port 8082. All processes read their database configuration from
the environment variables injected by the \texttt{http-diff-job},
which means they all point at the same fixture (typically
\texttt{DB\_3.0\_J} for correctness runs).

\paragraph{Phase 2, Deploy verification.}
Before any test case is sent, the harness polls
\texttt{GET /api/ping} on each active port until every running
instance returns \texttt{200 OK}, with a generous time-out. This
handshake was added in response to spurious early failures observed
during development, where the first request of a test would happen
to land on a still-deploying container; the corresponding commit
message in the repository history is the literal sentence
``\emph{ci: HttpDiff, prove deploy actually happened before
counting it as ready}''.

\paragraph{Phase 3, Test replay.}
The list of test cases lives in
\texttt{src/\allowbreak test/\allowbreak resources/\allowbreak http-diff-cases.json}, a small \ac{JSON}
array in which each entry describes an \ac{HTTP} request (method,
path, headers, optional body). Listing~\ref{lst:httpcases} shows the
shape of the file.

\begin{lstlisting}[caption={Excerpt from \texttt{http-diff-cases.json}, showing two of the test cases used by \texttt{HttpDiff}.},label=lst:httpcases]
[
  {
    "name": "ping",
    "method": "GET",
    "path": "/api/ping",
    "headers": { "Accept": "application/json" }
  },
  {
    "name": "station-short-json",
    "method": "GET",
    "path": "/api/stations/station/short/json",
    "headers": { "Accept": "application/json" }
  }
]
\end{lstlisting}

For each case, the harness fires the request against every active
port in parallel, captures status code, response headers and body,
and then performs a three-way comparison: \ac{HEAD} against the
recent baseline (if present) and \ac{HEAD} against the fixed
baseline (if present). The two comparisons are kept separate so that
``a change my commit introduced'' and ``a change that has accumulated
since the fixed reference'' remain distinguishable in the report.
Headers irrelevant to behavioural equivalence
(\texttt{Date}, \texttt{Server}, \texttt{Content-Length}) are
explicitly ignored. The body is compared as a normalised string,
sorted \ac{JSON} object keys, trimmed whitespace, in order to
avoid spurious diffs that would mask the real ones.

\paragraph{Phase 4, Tear-down and reporting.}
After the last case, all child processes are killed and the harness
emits two \ac{CSV}/\ac{JSON} reports, one per baseline, under
\texttt{target/http-diff-report/recent/} and
\texttt{target/http-diff-report/fixed/}. A combined JUnit XML report
under \texttt{target/surefire-reports/} flags each diff as a failed
test case, tagged with its baseline label, so that GitLab
shows the regressions both on the merge-request page and on the
pipeline page.

\section{Reports, Operations and Findings}
\label{sec:reports}

Each pipeline run publishes four artifact paths that survive for
between one and four weeks: the integration-test reports, the
validate-diff report, the \ac{HTTP}-diff report, and the JUnit XML
files consumed by GitLab itself. The retention windows are
deliberately asymmetric: integration test results expire faster
because they are cheap to reproduce, while diff reports are kept
longer because they are the historical record of behavioural change
during the refactor.

During the development of the framework the harnesses surfaced a
handful of real regressions that had been introduced by what looked
like cosmetic refactors. The most representative example is the
extraction of latitude and longitude range bounds to named
constants, which was reverted after the diff harness reported a
silent behavioural change in the boundary handling, the
corresponding commits in the repository history are
\texttt{7c8ff37 (refactor: extract lat/lon range bounds to named
constants)} and \texttt{cc223ba (Revert "refactor: extract lat/lon
range bounds to named constants")}. From the point of view of this
project, those two commits are the best validation that the
framework does the job it was designed for: it caught a regression
that no human reading of the diff was likely to catch by eye.

\section{Conclusion}
\label{chap5:sec:concs}

The application of the framework to \texttt{api\_v4.0} demonstrates
that a small, focused set of \ac{CI} jobs is sufficient to provide a
useful safety net during a long-running refactor of a Jakarta EE
\ac{API}, provided that the harnesses are designed around the
specifics of the system: a moving baseline, multiple response
formats, an internal database. The next chapter draws the main
conclusions of the project and outlines the items that, by choice
or by constraint, were left out of its scope.
