\chapter{Methodology and Architecture}
\label{chap:metodo-arq}

\section{Introduction}
\label{chap4:sec:intro}

This chapter describes the methodology by which the testing
infrastructure of the project was developed and then explains its
overall architecture. The methodology is openly iterative: rather
than building the final solution directly on top of the production
\ac{API}, two intermediate prototype applications were written first,
each addressing a separate concern in isolation, and the lessons
learnt from them were carried over into the production application.
Section~\ref{sec:legacy} situates the work against the legacy
\ac{API} that motivated the refactor. Section~\ref{sec:methodology}
describes the iterative methodology. Sections~\ref{sec:javaeeapp}
and~\ref{sec:syncspace} discuss the two prototypes,
\texttt{java-ee-app} and \texttt{syncspace-jakarta-app}. Finally,
Section~\ref{sec:arch} gives a high-level view of the architecture
of the differential testing framework that is detailed at the code
level in Chapter~\ref{chap:caso-epos}.

\section{The Legacy \ac{API} and the Need for a Safety Net}
\label{sec:legacy}

The starting point of the work is, to be blunt about it, a code base
that had grown organically for several years and that nobody fully
understood any more. The previous version of the \ac{EPOS} \ac{API}
contained Java methods of more than five thousand lines, source files
of more than thirty thousand lines, and many subtle behavioural
quirks introduced as patches by successive developers in response to
specific bug reports. The corollary was that no member of the
laboratory could realistically guarantee, for an arbitrary change,
that no client of the \ac{API} would be silently broken. Manual
testing covered a few well-known requests and tacitly relied on the
expectation that nobody would do anything too creative.

The full rewrite of the \ac{API} on top of Jakarta EE 11 made the
internal structure much cleaner, short methods, well-bounded
\ac{JAX-RS} resource classes, clear separation between parsing,
validation, query construction and serialization, but it also
opened a window of risk: at any given commit during the refactor, an
endpoint may produce subtly different output from the version
deployed in production. The role of this project is to build the
safety net that turns those subtle differences into a visible,
investigable signal in the \ac{CI} pipeline.

\section{Iterative Methodology}
\label{sec:methodology}

The work was organised into three iterations, each carried out in its
own dedicated repository. This separation served two purposes. First,
it kept each iteration small and focused: the first prototype is
deliberately stripped down to two trivial endpoints, so that the
entire complexity is in the \ac{CI} configuration and not in the
application; the second prototype is the opposite, focusing on
application code and leaving \ac{CI} aside; the production \ac{API}
then brings the two strands back together. Second, the prototypes
have value in their own right as didactic artifacts that future
students of the laboratory can read and re-use.

The three iterations are summarised in Table~\ref{tab:iterations}.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{p{3.0cm} p{4.5cm} p{4.5cm}}
    \toprule
    \textbf{Iteration} & \textbf{Focus} & \textbf{What was carried over} \\
    \midrule
    Prototype 1 \newline \texttt{java-ee-app}
      & \ac{CI}/\ac{CD} pipeline patterns: dual Payara, Arquillian,
        JUnit XML reporting.
      & The shape of the \texttt{.gitlab-ci.yml}, the Maven cache
        strategy and the choice of base Docker image. \\
    \addlinespace
    Prototype 2 \newline \texttt{syncspace-jakarta-app}
      & Jakarta EE 11 features: \ac{CDI}, \ac{JAX-RS}, \ac{JPA},
        Bean Validation, MicroProfile JWT, exception mappers.
      & Validation patterns, exception mapper layout, record-based
        DTOs, some of which were adopted by the refactor of the
        production \ac{API}, some only used didactically. \\
    \addlinespace
    Production \newline \texttt{api\_v4.0}
      & End-to-end integration: full pipeline, code-level diff,
        \ac{HTTP}-level diff against real database via \ac{VPN}.
      & Final artifact of the project. \\
    \bottomrule
  \end{tabular}
  \caption{The three iterations of the project and their respective
  focus.}
  \label{tab:iterations}
\end{table}

\section{Prototype 1: \texttt{java-ee-app}}
\label{sec:javaeeapp}

The first prototype is a deliberately minimal Jakarta EE
application whose role is to validate the \ac{CI}/\ac{CD} pipeline.
It exposes a single \ac{JAX-RS} application class with two trivial
endpoints, a \texttt{Hello World} text endpoint and a small
\ac{POJO} serialised to \ac{JSON}, and a pair of Arquillian
integration tests that exercise them against a managed Payara
container.

\subsection{Layout and Stack}

The application is a Maven \ac{WAR} project targeting Java SE 17 and
Jakarta EE 11. Its dependencies, beyond the Jakarta EE \acs{API}
bundle (declared with \texttt{provided} scope), are JUnit 5
Jupiter and the Arquillian \texttt{payara-server-managed} adapter.
The \texttt{src/test} folder contains an \texttt{AbstractBaseIT.java}
that prepares a ShrinkWrap deployment and a \texttt{JsonApplicationIT.java}
that issues \ac{HTTP} requests against the deployed application.

\subsection{The \ac{CI} Pipeline}

The interesting part of this prototype is the \texttt{.gitlab-ci.yml},
shown in Listing~\ref{lst:javaeeapp-ci}. The pipeline has two
stages: \texttt{build}, which produces the \ac{WAR}, and
\texttt{test}, which runs the integration tests against
\emph{two} Payara Server containers, declared as
\emph{services} in the job definition. The dual-server setup was
useful as a stress test of the runner network: if the pipeline can
boot two containers and run tests against them, it can certainly boot
the single Payara Micro instance used by the production pipeline.

\begin{lstlisting}[language=yaml,caption={\texttt{.gitlab-ci.yml} of the first prototype, two-stage pipeline with dual Payara services.},label=lst:javaeeapp-ci]
default:
  image: maven:3.9.6-eclipse-temurin-21
  tags:
    - ubi-runner

cache:
  key: global-maven-cache
  paths:
    - .m2/repository

variables:
  MAVEN_OPTS: "-Dmaven.repo.local=$CI_PROJECT_DIR/.m2/repository"

stages:
  - build
  - test

build-job:
  stage: build
  script:
    - mvn clean package -DskipTests
  artifacts:
    paths:
      - target/*.war

integration-test-job:
  stage: test
  services:
    - name: payara/server-full:7.2026.5
      alias: payara-node-1
    - name: payara/server-full:7.2026.5
      alias: payara-node-2
  script:
    - sleep 45
    - mvn verify
  artifacts:
    when: always
    reports:
      junit:
        - "target/failsafe-reports/failsafe-summary.xml"
        - "target/failsafe-reports/TEST-*.xml"
    paths:
      - target/failsafe-reports/
    expire_in: 1 week
\end{lstlisting}

Three patterns from this prototype were carried over to the
production pipeline:

\begin{enumerate}
  \item the pinned base image and the \texttt{ubi-runner} tag, which
        together guarantee that the job runs on a known runner with a
        known toolchain;
  \item the global Maven cache, which avoids re-downloading the
        repository on every job;
  \item the \texttt{reports: junit:} stanza, which is what makes
        GitLab understand failed assertions as failed pipeline steps
        rather than as opaque exit codes.
\end{enumerate}

The forty-five-second \texttt{sleep} that precedes \texttt{mvn
verify} compensates for the time required by the Payara containers
to become reachable; a cleaner health-poll loop was tried and
discarded as not worth the additional complexity for a prototype.

\section{Prototype 2: \texttt{syncspace-jakarta-app}}
\label{sec:syncspace}

The second prototype goes in the opposite direction. Where the first
prototype is a few lines of application code with a complete \ac{CI}
pipeline, the second prototype is a complete application with no
\ac{CI} pipeline at all. Its goal is to exercise the Jakarta EE 11
specifications listed in Section~\ref{sec:java} on a domain that is
small enough to fit in a student's head but large enough to require
non-trivial use of each specification.

\subsection{The Domain}

\texttt{syncspace-jakarta-app} is the back end of a hypothetical
room-booking service for a co-working space. The domain is centred
on three entities: \texttt{User}, \texttt{Room} and \texttt{Booking}.
A room has a type (\texttt{HOT\_DESK}, \texttt{PODCAST\_STUDIO},
\texttt{BOARDROOM}, \texttt{EVENT\_SPACE}), a maximum capacity and
an active flag. A booking links a user to a room over a time
interval and counts the number of attendees. Users authenticate
through a login endpoint that issues a \ac{JWT}; subsequent calls to
the booking endpoints are protected by MicroProfile
\ac{JWT}~\cite{microprofile}.

\subsection{Jakarta Specifications Exercised}

Each part of the application was deliberately written to use a
specific Jakarta or MicroProfile specification:

\begin{itemize}
  \item \textbf{\ac{JAX-RS} 4.0~\cite{jaxrs}.} The
        \texttt{SyncSpaceApplication} class declares an
        \texttt{@ApplicationPath("/api/v1")} and the resources
        (\texttt{AuthResource}, \texttt{BookingResource},
        \texttt{RoomResource}) follow standard \ac{JAX-RS} idioms,
        with \texttt{@Inject} for service dependencies and the
        \texttt{Response} builder for fluent response construction.
  \item \textbf{\ac{CDI} 4.1.} Services and exception mappers are
        \ac{CDI} beans, injected into the \ac{JAX-RS} resources.
        \texttt{AuthResource} is \texttt{@RequestScoped}; services
        are implicitly singletons.
  \item \textbf{\ac{JPA} 3.2~\cite{jpa}.} Entities use
        \texttt{@Entity}, \texttt{@ManyToOne}, \texttt{@OneToMany},
        \texttt{@Enumerated(STRING)} for the enums, and
        \texttt{@PrePersist} for default timestamps.
  \item \textbf{Bean Validation 3.1.} A custom
        \texttt{@ValidDateRange} class-level annotation is paired
        with a validator that ensures \texttt{endTime} is strictly
        greater than \texttt{startTime} and attaches the error to
        the correct property through
        \texttt{addPropertyNode("endTime")}.
  \item \textbf{MicroProfile JWT 2.1.} The
        \texttt{@LoginConfig(authMethod = "MP-JWT")} on the
        application class enables token-based authentication; tokens
        are signed manually with Nimbus JOSE+JWT using an RSA
        signature, with the user identifier as the subject and the
        user's roles as the \texttt{groups} claim.
  \item \textbf{Exception mappers.} Four dedicated mappers
        (\texttt{BusinessValidationMapper}, \texttt{ValidationExceptionMapper},
        \texttt{JsonProcessingExceptionMapper},
        \texttt{GlobalExceptionMapper}) translate domain and
        framework exceptions into structured \ac{JSON} responses,
        avoiding the leaking of internal stack traces to the
        \ac{API} consumer. The end-to-end flow is summarised in
        Figure~\ref{fig:errormapper}.
\end{itemize}

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\small},
      node distance=8mm,
      box/.style={draw, rounded corners=2pt, align=center, thick,
                  minimum height=8mm, minimum width=28mm},
      stage/.style={box, fill=dkblue!10},
      mapper/.style={box, fill=mauve!10, minimum width=36mm, font=\footnotesize},
      outbox/.style={box, fill=dkgreen!10, minimum width=36mm},
      arrow/.style={-{Stealth[length=2mm]}, thick}
    ]
    \node[stage] (client) {HTTP client};
    \node[stage, below=of client] (resource) {\textbf{BookingResource} \\\footnotesize \texttt{@Inject BookingService}};
    \node[stage, below=of resource] (svc) {\textbf{BookingService} \\\footnotesize business logic};
    \node[stage, below=of svc, fill=red!12] (ex) {Exception thrown};
    \node[stage, below=of ex] (dispatch) {JAX-RS exception dispatch \\\footnotesize matches type};

    \node[mapper, below=10mm of dispatch, xshift=-58mm] (m1) {BusinessValidation\\Mapper};
    \node[mapper, right=4mm of m1] (m2) {Validation\\ExceptionMapper};
    \node[mapper, right=4mm of m2] (m3) {JsonProcessing\\ExceptionMapper};
    \node[mapper, right=4mm of m3] (m4) {Global\\ExceptionMapper};

    \node[outbox, below=10mm of m2, xshift=20mm] (resp) {Response.builder() \\\footnotesize \texttt{\{ "error": "...", "details": [...] \}}};
    \node[stage, below=of resp] (cliback) {HTTP client \\\footnotesize sees structured JSON, never a stack trace};

    \draw[arrow] (client)   -- node[right,font=\scriptsize]{POST /api/v1/bookings} (resource);
    \draw[arrow] (resource) -- (svc);
    \draw[arrow] (svc)      -- (ex);
    \draw[arrow] (ex)       -- (dispatch);
    \draw[arrow] (dispatch) -- (m1);
    \draw[arrow] (dispatch) -- (m2);
    \draw[arrow] (dispatch) -- (m3);
    \draw[arrow] (dispatch) -- (m4);
    \draw[arrow] (m1)       -- (resp);
    \draw[arrow] (m2)       -- (resp);
    \draw[arrow] (m3)       -- (resp);
    \draw[arrow] (m4)       -- (resp);
    \draw[arrow] (resp)     -- (cliback);
  \end{tikzpicture}%
  }
  \caption{The \ac{CDI} + \ac{JAX-RS} exception-handling flow used by
  \texttt{syncspace-jakarta-app} and partly inherited by
  \texttt{api\_v4.0}. Any exception thrown inside a resource or
  service bean is intercepted by the most specific registered
  \texttt{ExceptionMapper}, which produces a stable \ac{JSON}
  error body.}
  \label{fig:errormapper}
\end{figure}

\subsection{What was Carried Over}

Most of the code of this prototype was never reused in the
production \ac{API}, the production domain is very different
from a room booker. Two patterns were nevertheless adopted into the
refactor. The first is the use of a small set of focused
\ac{JAX-RS} exception mappers rather than a single catch-all
\ac{JAX-RS} \texttt{ExceptionMapper<Throwable>}: this leaves
\ac{HTTP} status codes meaningful and gives the differential test
harness a stable signal to assert on. The second is the use of
\texttt{record}-based \acp{POJO} for request and response bodies,
which avoids hand-written \texttt{equals}/\texttt{hashCode} and
makes diffing two responses simpler.

The remaining features, in particular the full \ac{JWT} flow,
were retained in the prototype as a self-contained reference
example to which future students can look for guidance on Jakarta
EE 11 idioms.

\section{Architecture of the Differential Testing Framework}
\label{sec:arch}

The architecture of the final testing framework, applied to the
production \ac{API}, is summarised in Figure~\ref{fig:arch}. It is
made of three concentric layers: the GitLab pipeline, the harness
processes, and the application instances themselves.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      node distance=8mm and 12mm,
      every node/.style={font=\small},
      box/.style={draw, rounded corners=2pt, minimum height=10mm,
                  minimum width=28mm, align=center, thick},
      smallbox/.style={box, minimum width=22mm, minimum height=8mm,
                  font=\footnotesize},
      arrow/.style={-{Stealth[length=2mm]}, thick}
    ]
    \node[box, fill=lgray] (push) {Developer push \\ on a branch};
    \node[box, right=18mm of push, fill=dkblue!10] (gitlab) {GitLab CI server \\ (self-managed)};
    \node[box, right=18mm of gitlab, fill=dkblue!10] (runner) {GitLab Runner \\ (\texttt{ubi-runner-.31})};

    \node[smallbox, below=12mm of runner, fill=dkgreen!10] (build) {\texttt{build-job} \\ \scriptsize mvn package};
    \node[smallbox, below=of build, fill=dkgreen!10] (test) {\texttt{integration-test-job}};
    \node[smallbox, below=of test, fill=dkgreen!10] (diff)  {\texttt{validate-diff-job}};
    \node[smallbox, below=of diff, fill=dkgreen!10] (http)  {\texttt{http-diff-job}};

    \node[box, left=24mm of test, fill=mauve!10] (head) {HEAD WAR};
    \node[box, below=of head, fill=mauve!10] (brecent) {Baseline WAR \\ recent \\ (\texttt{HEAD\textasciitilde 1})};
    \node[box, below=of brecent, fill=mauve!10] (bfixed) {Baseline WAR \\ fixed \\ (\texttt{\$DIFF\_FIXED\_BASELINE})};

    \node[box, below=10mm of bfixed, fill=dark!10] (db) {PostgreSQL \\ (\texttt{192.168.168.30}) \\ \scriptsize J or F fixture};

    \draw[arrow] (push) -- (gitlab);
    \draw[arrow] (gitlab) -- (runner);
    \draw[arrow] (runner) -- (build);
    \draw[arrow] (build)  -- (test);
    \draw[arrow] (test)   -- (diff);
    \draw[arrow] (diff)   -- (http);

    \draw[arrow, dashed] (build) -- (head);
    \draw[arrow, dashed] (diff)  -- (brecent);
    \draw[arrow, dashed] (diff)  -- (bfixed);
    \draw[arrow, dashed] (http)  -- (head);
    \draw[arrow, dashed] (http)  -- (brecent);
    \draw[arrow, dashed] (http)  -- (bfixed);
    \draw[arrow, dashed] (head)    -- (db);
    \draw[arrow, dashed] (brecent) -- (db);
    \draw[arrow, dashed] (bfixed)  -- (db);
  \end{tikzpicture}%
  }
  \caption{Overall architecture of the differential testing framework.
  Solid arrows denote control flow inside the pipeline; dashed arrows
  denote artifact or data access. Both diff jobs run a three-way
  comparison between the \ac{HEAD} build and two baselines, the
  immediate parent commit and a fixed reference commit, all
  pointed at the same database fixture.}
  \label{fig:arch}
\end{figure}

The flow is as follows. A push by the developer is received by the
self-managed GitLab server, which schedules a pipeline on the runner
tagged \texttt{ubi-runner-.31}. The pipeline goes through four jobs
across three stages: \texttt{build-job} produces the \ac{HEAD}
\ac{WAR}; \texttt{integration-test-job} runs the regular JUnit
integration tests against that \ac{WAR}; \texttt{validate-diff-job}
reconstructs both baseline \acp{WAR} (one from
\texttt{HEAD\textasciitilde 1}, one from the fixed commit in
\texttt{\$DIFF\_FIXED\_BASELINE}) using Git worktrees and runs the
code-level differential harness against each in turn;
\texttt{http-diff-job} boots Payara Micro instances of all three
\acp{WAR} on different ports, connects them to the real database
over the laboratory's network and replays the test cases listed in
\texttt{http-diff-cases.json} against all of them, collecting
per-case diffs against both baselines independently. The two diff jobs publish reports that
GitLab understands both as JUnit-style results and as raw
artifacts for later inspection.

\section{Conclusion}
\label{chap4:sec:concs}

The iterative methodology adopted in this project, two prototypes
followed by the production application, traded an apparently
longer time-to-result for a much sharper separation of concerns
between \ac{CI}/\ac{CD} infrastructure and Jakarta EE feature usage.
The first prototype gave the \texttt{.gitlab-ci.yml} its shape; the
second prototype gave the production refactor a small but useful
set of Jakarta EE idioms. The next chapter applies the resulting
framework to the production \ac{API} and walks through its
\ac{CI} pipeline, the two diff harnesses and their reports.
