\chapter{Technologies and Tools Used}
\label{chap:tecno-ferra}

\section{Introduction}
\label{chap3:sec:intro}

This chapter describes the concrete set of technologies and tools on
which the work reported in this document is built. The set was not
chosen in a single decision, but converged over the course of the
project through a mixture of constraints (what the production
\ac{API} already uses, what the laboratory's network allows, what the
proposal explicitly mandates), comparisons (between equivalent
alternatives) and pragmatic preference. The chapter is structured by
layer of the stack: \ac{API} contracts and serialization formats in
Section~\ref{sec:apis}, the Java/Jakarta ecosystem in
Section~\ref{sec:java}, persistence and \ac{SQL} databases in
Section~\ref{sec:sql}, application and \ac{CI} servers in
Section~\ref{sec:servers}, version control in
Section~\ref{sec:gitsec}, containerization in Section~\ref{sec:docker}
and network access in Section~\ref{sec:vpn}.

\section{APIs, Serialization Formats and Specification}
\label{sec:apis}

\subsection{REST and the Multi-format Response}

All endpoints exposed by the applications discussed in this report
follow the \ac{REST} architectural style as described by
Fielding~\cite{fielding2000rest}: resources are identified by stable
\acp{URL}, manipulated through a small set of \ac{HTTP} verbs (with
\texttt{GET} dominating for read-only \ac{GNSS} metadata), and the
\ac{API} is stateless on the server side. There is no session
affinity, which is a property that the differential test harness
described in Chapter~\ref{chap:caso-epos} relies on heavily: the same
request can be replayed against two independent server processes and
the responses can be compared directly.

The production \ac{API} (\texttt{api\_v4.0}) is unusual in that it
serves the \emph{same} resource in several serialization formats,
selected by a path parameter at the end of the \ac{URL}. The supported
formats are \ac{JSON} (default), \ac{XML} (produced through Jackson's
XML data format module), \ac{CSV} (produced through a custom writer)
and, for the file metadata endpoints, an executable shell
\emph{script} that downloads the corresponding files. This multi-format
output was inherited from the legacy \ac{API}, many of its clients
are scientific scripts that consume \ac{CSV} or \ac{JSON} directly,
and was preserved in the refactor; it is also the reason the
\ac{HTTP}-level differential harness has to compare not only headers
and status codes but also the body, format by format.
Figure~\ref{fig:multiformat} sketches the content negotiation pattern
that turns a single resource method into four wire formats.

\begin{figure}[h]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
      every node/.style={font=\small},
      box/.style={draw, rounded corners=2pt, align=center, thick,
                  minimum height=9mm, minimum width=22mm},
      writer/.style={box, fill=dkgreen!10, minimum width=20mm},
      arrow/.style={-{Stealth[length=2mm]}, thick}
    ]
    \node[box, fill=dkblue!10] (req) {\texttt{GET /api/stations/.../\{format\}}};
    \node[box, fill=mauve!10, below=8mm of req, minimum width=46mm]
      (resource) {\textbf{StationEndpoints}\\[2pt]\footnotesize same Java method, regardless of format};
    \node[box, below=8mm of resource, minimum width=46mm, fill=lgray]
      (negot) {Content negotiation \\\footnotesize (path parameter \texttt{\{format\}})};

    \node[writer, below=10mm of negot, xshift=-36mm] (json)   {\textbf{JSON} \\\footnotesize Jackson};
    \node[writer, right=4mm of json]                 (xml)    {\textbf{XML}  \\\footnotesize jackson-xml};
    \node[writer, right=4mm of xml]                  (csv)    {\textbf{CSV}  \\\footnotesize custom writer};
    \node[writer, right=4mm of csv]                  (script) {\textbf{Script}\\\footnotesize files only};

    \node[box, below=10mm of csv, fill=dark!10, minimum width=66mm]
      (resp) {HTTP response body};

    \draw[arrow] (req)       -- (resource);
    \draw[arrow] (resource)  -- (negot);
    \draw[arrow] (negot)     -- (json);
    \draw[arrow] (negot)     -- (xml);
    \draw[arrow] (negot)     -- (csv);
    \draw[arrow] (negot)     -- (script);
    \draw[arrow] (json)      -- (resp);
    \draw[arrow] (xml)       -- (resp);
    \draw[arrow] (csv)       -- (resp);
    \draw[arrow] (script)    -- (resp);
  \end{tikzpicture}%
  }
  \caption{Multi-format content negotiation in \texttt{api\_v4.0}. The
  same Java resource method is reused across four serializations,
  selected by a path parameter, and the differential harness must
  compare the body of each format separately.}
  \label{fig:multiformat}
\end{figure}

\subsection{\ac{JSON} and \ac{XML}}

\ac{JSON} is the default serialization and the format on which most
of the test fixtures of this project rely. The Jakarta \ac{REST}
implementation provided by Payara takes care of object serialization
through Jackson, with negligible per-request overhead. \ac{XML} is
provided by the \texttt{jackson-dataformat-xml} module, configured to
preserve the same Java models that produce \ac{JSON}. This dual
configuration is valuable because it removes a whole class of
\emph{schema drift} bugs: as long as the Java model is correct, both
\ac{JSON} and \ac{XML} outputs are consistent with each other.

\subsection{\ac{CSV}}

\ac{CSV} is the format favoured by historical clients of the \ac{API}
and is produced through a small set of writer classes that flatten
the entity tree into rows. The differential harness handles \ac{CSV}
exactly the same way it handles \ac{JSON} and \ac{XML}: the textual
body of the response is treated as an opaque blob that must match,
byte for byte, against the baseline.

\subsection{\ac{OpenAPI} and Swagger \ac{UI}}

The contract of the production \ac{API} is described in an
\ac{OpenAPI} 3.1.0~\cite{openapi} document
(\texttt{openapi.json}) that ships inside the \ac{WAR}. The same
document is served to a Swagger \ac{UI}~\cite{swaggerui} instance
hosted at \texttt{/swagger.html}, which provides an interactive,
in-browser playground for the endpoints. Two practical benefits arise
from this setup. First, the document is the single source of truth
shared with downstream consumers of the \ac{API}, in particular the
\ac{EPOS} portal. Second, the same document can be used to drive
contract tests: a future evolution of the test framework, briefly
discussed in Chapter~\ref{chap:conc-trab-futuro}, is to derive the
\ac{HTTP} differential test cases automatically from the \ac{OpenAPI}
specification instead of maintaining them in a separate \ac{JSON} file.

\section{Java and Jakarta EE}
\label{sec:java}

\subsection{The Java Platform}

All applications discussed in this report run on the Java Platform.
The production \ac{API} and the second prototype target Java SE
21~\cite{javaSE21}, the long-term support release at the time of the
project, while the first prototype is built against Java SE 17,
the version against which its base \ac{CI} image (\texttt{maven:3.9.6-eclipse-temurin-21})
remains backwards-compatible. The choice of Java was effectively
imposed by the existing code base, but is independently justified by
the maturity of the platform, the strength of its enterprise
ecosystem and the very wide set of tools (\acp{IDE}, build systems,
test frameworks) that target it.

The Java records introduced in recent versions of the language were
used liberally in the prototypes for immutable request and response
\acp{POJO}, and the new \texttt{switch} expression syntax is used in
several places where the refactored validation logic replaced long
\texttt{if/else} chains from the legacy \ac{API}.

\subsection{Jakarta EE and the Move from Java EE}

Java EE was renamed to Jakarta EE in 2018, when its stewardship moved
from Oracle to the Eclipse Foundation~\cite{jakartaee}. The most
visible technical consequence of the move was the change of the root
package from \texttt{javax.*} to \texttt{jakarta.*}, starting with
Jakarta EE 9. For an application of the size of the production
\ac{API}, this was not a trivial change, but it had already been
absorbed by the time this project started; the current code base
targets Jakarta EE 11.

The specific Jakarta specifications used in this work are
summarised in Table~\ref{tab:jakarta-specs}.

\begin{table}[h]
  \centering
  \small
  \begin{tabular}{l p{6.5cm} l}
    \toprule
    \textbf{Specification} & \textbf{Purpose} & \textbf{Version} \\
    \midrule
    Jakarta \ac{REST} (\ac{JAX-RS}) & HTTP endpoints, content negotiation
       & 4.0 \\
    Jakarta \ac{CDI} & Dependency injection, bean lifecycle
       & 4.1 \\
    Jakarta Persistence (\ac{JPA}) & Object-relational mapping
       & 3.2 \\
    Jakarta Bean Validation & Declarative constraint checking
       & 3.1 \\
    Jakarta JSON Binding (JSON-B) & POJO $\leftrightarrow$ \ac{JSON}
       & 3.0 \\
    Eclipse MicroProfile JWT & JSON Web Token authentication
       & 2.1 \\
    Eclipse MicroProfile Health & Liveness/readiness probes
       & 4.0 \\
    \bottomrule
  \end{tabular}
  \caption{Jakarta EE and MicroProfile specifications used across the
  three applications.}
  \label{tab:jakarta-specs}
\end{table}

\subsection{Why Jakarta EE rather than Spring Boot}

A common alternative to Jakarta EE for a green-field Java \ac{API} is
Spring Boot~\cite{springBoot}. The two are largely equivalent at the
level of features, but they differ in philosophy: Spring Boot bundles
an embedded servlet container and a curated, opinionated
\ac{BOM}; Jakarta EE relies on an external application server, in
this case Payara, which provides the runtime contracts that the
application targets.

For the prototypes, Spring Boot would have been a perfectly viable
choice. For the production \ac{API}, however, the existing code base
is already on Jakarta EE and is deployed onto an existing Payara
server inside the laboratory's network. Migrating to Spring Boot
would have meant a second, parallel refactor on top of the one already
in progress, with no proportional benefit. Sticking with Jakarta EE
was therefore the lower-risk option.

\section{\ac{SQL} and the Database Layer}
\label{sec:sql}

\subsection{\ac{SQL} as a Lingua Franca}

The persistence model of the production \ac{API} is heavily relational
and the queries themselves are non-trivial: \ac{GNSS} stations,
networks, agencies, countries, file types, observation summaries and
embargo windows are all linked through foreign keys, and most
endpoints execute several joins per request. \ac{SQL}~\cite{postgresql}
is therefore the natural query language for this domain; even where
the application code uses \ac{JPA}'s object-oriented abstractions
(Section~\ref{sec:jpa-hibernate}), what reaches the database is plain
\ac{SQL}.

\subsection{PostgreSQL and MySQL}

The laboratory standardised on \textbf{PostgreSQL}~\cite{postgresql}
for the \ac{EPOS} production database. PostgreSQL was preferred over
\textbf{MySQL}~\cite{mysql} for two main reasons. First, PostgreSQL's
adherence to the \ac{SQL} standard, in particular for advanced
\texttt{JOIN} types, common table expressions, the \texttt{numeric}
type used for the high-precision station coordinates, and the rich
set of available index types (B-tree, BRIN, GiST and others),
made the rewrite of the legacy queries simpler. Second, PostgreSQL
leaves the door open to the PostGIS extension for the bounding-box
queries that the \ac{GNSS} domain naturally invites; that extension
is not in use today (see Section~\ref{sec:epos-db}) but is a
realistic future addition. MySQL was nevertheless used in the
docker-compose configuration of the second prototype
(Section~\ref{sec:syncspace}) to validate that the application-level
code is not tied to a specific dialect.

\subsection{\ac{JPA}, Hibernate and the Persistence Unit}
\label{sec:jpa-hibernate}

\ac{JPA}~\cite{jpa} is the Jakarta specification for object-relational
mapping in Java. The production \ac{API} uses
Hibernate~\cite{hibernate} as the \ac{JPA} provider, configured
through a \texttt{persistence.xml} file. The persistence unit is
named \texttt{MeuPU} and is declared as
\texttt{RESOURCE\_LOCAL}, which means that transactions are managed
explicitly by application code rather than by the container, a
deliberate choice that keeps the differential test harness in full
control of when transactions begin and end.

Database connections are pooled through \textbf{HikariCP}~\cite{hikaricp}
with a small, conservative pool (2--15 connections, 20-second connect
timeout, 30-second idle, 5-second validation), tuned for the
read-mostly workload of the \ac{API} and for the relatively low
concurrency observed in production.

\subsection{The \ac{EPOS} Database in Practice}
\label{sec:epos-db}

A closer look at the production database is useful at this point,
because most of the discussion in the following chapters takes its
shape and its peculiarities for granted. The schema lives in the
\texttt{public} schema of a PostgreSQL 16 instance and is shared by
two database fixtures, both reachable from the laboratory's internal
network on port 5432 of \texttt{192.168.168.30}:
\texttt{DB\_3.0\_J} hosts a small but representative production
subset; \texttt{DB\_3.0\_F} hosts the full dataset and is reserved
for volume-oriented and performance experiments.

\paragraph{Schema shape.}
The current schema contains forty-two tables and roughly fifty
foreign-key constraints; the model is highly normalised, with no
table currently exceeding thirty columns. The tables fall into four
broad families:

\begin{itemize}
  \item \emph{Core domain}, \texttt{station}, \texttt{network},
        \texttt{agency}, \texttt{country}, \texttt{rinex\_file} and
        \texttt{highrate\_rinex\_file}. These hold the entities that
        appear in the public \ac{API} payloads.
  \item \emph{Junction tables}, \texttt{agency\_network},
        \texttt{station\_network}, \texttt{station\_agency\_role},
        \texttt{node\_station} and \texttt{node\_data\_center}.
        These materialise the many-to-many relationships that the
        domain calls for (a station belongs to one or more networks,
        an agency operates one or more networks, and so on).
  \item \emph{Quality-control summaries}, a family of seven tables
        (\texttt{qc\_report\_summary}, \texttt{qc\_constellation\_\-summary},
        \texttt{qc\_header\_observations}, \texttt{qc\_navigation\_msg}
        and four typed \texttt{qc\_observation\_summary\_\{c,d,l,s\}}
        variants for code, Doppler, phase-locked and signal-strength
        observations). These store the post-processing of each
        \ac{RINEX} file.
  \item \emph{Auxiliary}, embargoes, documents, licenses, logs,
        and the node/data-centre infrastructure tables; together with
        a small \texttt{authentication} table used by the legacy
        admin endpoints.
\end{itemize}

\paragraph{One small but instructive design choice.}
The \texttt{country} table uses the ISO three-letter code
(\texttt{iso\_code}) as its natural primary key instead of a surrogate
\texttt{id}, and \texttt{station.country\_code} is a direct
\texttt{character(3)} foreign key onto it. This single decision
removes the most common ``which agency are you talking about'' join
ambiguity from station queries and is a representative example of the
small simplifications the schema makes possible.

\paragraph{Spatial data: today and tomorrow.}
Station coordinates are stored as plain \texttt{numeric(12,3)} for
the Cartesian \texttt{x/y/z} components and as \texttt{numeric(6,4)}
/ \texttt{numeric(7,4)} for the geodetic \texttt{lat}/\texttt{lon}.
The PostGIS extension is \emph{not} currently installed, the only
non-default extension is \texttt{plpgsql}. Bounding-box and polygon
checks are implemented in Java by the helpers of the \texttt{Geometry}
package (Section~\ref{sec:apiv4-arch}), which is sufficient for
present needs and keeps the database side neutral. Migrating those
checks to PostGIS is one of the items in the project's future-work
list (Chapter~\ref{chap:conc-trab-futuro}), but is expected to make
little difference to correctness; the motivation, when it comes, will
be query plan quality on large datasets.

\paragraph{Indexing today and indexing tomorrow.}
The schema today carries about twenty-five explicit indexes, almost
all of them B-tree indexes on foreign-key columns
(\texttt{idx\_file} on \texttt{rinex\_file(id\_station)},
\texttt{idx\_qc\_filesum} on \texttt{qc\_report\_summary(id\_rinexfile)}
with \texttt{fillfactor=90} to favour reads over writes, and so on).
This handles the common access patterns of the \ac{API}, where every
endpoint joins from the \texttt{station} or \texttt{rinex\_file} side
through a chain of foreign keys.

The natural next step, and the main reason the larger
\texttt{DB\_3.0\_F} fixture was provisioned, is to validate the
introduction of \textbf{BRIN} (Block Range Index) indexes on the
date columns of \texttt{rinex\_file} (\texttt{reference\_date},
\texttt{published\_date}, \texttt{release\_date}) and of the
\texttt{highrate} variant. BRIN indexes are appropriate here because
\ac{RINEX} files are inserted in approximate date order, so the
on-disk row layout already correlates with the indexed column, which
is the precondition under which a BRIN index is small and fast. The
differential test framework introduced in Chapter~\ref{chap:caso-epos}
plays a role in this evaluation: by running the same diff cases
against the \texttt{DB\_3.0\_F} fixture before and after the BRIN
addition, the framework gives a concrete behavioural signal in
addition to whatever \texttt{EXPLAIN ANALYZE} would say in isolation.

\section{Application and \ac{CI} Servers}
\label{sec:servers}

\subsection{Payara Server}

\textbf{Payara Server}~\cite{payara} is a derivative of the GlassFish
reference implementation of Jakarta EE, maintained as a commercial
distribution by Payara Services Ltd. The laboratory adopted Payara
several years ago and the production \ac{API} is deployed onto an
existing Payara installation; this project did not change that
choice. Two specific properties of Payara matter for the work
reported here. First, Payara Server packages a full Jakarta EE
runtime, which is what the production \ac{WAR} is built against.
Second, Payara distributes a companion product, \textbf{Payara
Micro}, which is a single, self-contained \ac{JAR} that can be
launched from the command line with a \ac{WAR} on its argument list:
\begin{verbatim}
java -jar payara-micro.jar --port 8081 target/ROOT.war
\end{verbatim}
Payara Micro is the engine on which the \ac{HTTP}-level differential
harness relies: the harness boots two instances of Payara Micro on
two different ports, one serving the baseline \ac{WAR} and the other
the \ac{HEAD} \ac{WAR}, and replays the same requests against both.

\subsection{Apache Tomcat as an Alternative}

\textbf{Apache Tomcat}~\cite{apacheTomcat} is the most widely used
servlet container in the Java world, and is sometimes used as a
lightweight alternative to a full Jakarta EE server. It was
considered for the second prototype but ruled out because, unlike
Payara, Tomcat does not ship a full Jakarta EE 11 runtime,
\ac{CDI}, \ac{JPA} and Bean Validation would have had to be wired
in by hand. Standardising on Payara across the three projects
eliminated that overhead.

\subsection{Maven as Build Tool}

All three applications are built with \textbf{Apache Maven}.
The \texttt{pom.xml} of the production \ac{API} configures three
plugins worth highlighting: \texttt{maven-war-plugin}, which produces
the deployable \texttt{ROOT.war}; \texttt{maven-failsafe-plugin},
which runs integration tests (named \texttt{*IT.java}) during the
\texttt{verify} phase and emits \ac{JUnit} XML reports; and
\texttt{maven-dependency-plugin}, which downloads the Payara Micro
\ac{JAR} into \texttt{target/payara-micro/} during the build so that
the \ac{HTTP} harness can later spawn it as a subprocess.

\subsection{GitLab and GitLab Runner}

\textbf{GitLab}~\cite{gitlabci} was already discussed in
Chapter~\ref{chap:estado-da-arte} as the \ac{CI}/\ac{CD} platform
adopted by the laboratory. From the point of view of the technology
stack, two components are relevant. The \emph{GitLab server} is a
self-managed installation~\cite{gitlabSelfManaged} running on a
\ac{VM} inside the laboratory; it hosts the repository, the issues,
the pipeline definitions and the merge requests. The \emph{GitLab
Runner}~\cite{gitlabRunner} is a separate process, hosted on another
\ac{VM} (tagged \texttt{ubi-runner-.31}), that polls the GitLab
server for pending jobs and executes them inside a Docker container.
That separation is significant: it means the runner machine is the
one that needs network access to the laboratory's database, not the
GitLab server itself.

\section{Version Control: Git and GitLab}
\label{sec:gitsec}

\textbf{Git}~\cite{chacon2014pro} is the version control system used
across all three applications. Beyond the obvious use as a source
code repository, Git is used in two non-trivial ways in this project.
First, the \ac{CI} pipeline of the production \ac{API} uses
\texttt{git worktree} to materialise an additional copy of the
repository pinned to a baseline commit, on the runner's local file
system; this is the mechanism by which the differential harness
obtains the ``previous version'' against which the current build is
compared. Second, the harness identifies the baseline as the first
parent of \texttt{HEAD} (\texttt{HEAD\textasciitilde 1}), which makes
the comparison meaningful even on long-running refactor branches in
which the developer commits frequently. The exact pattern is
discussed in detail in Section~\ref{sec:pipeline}.

GitLab itself plays the role of the centralised Git server. All three
repositories are hosted under \texttt{gitlab.com}: the production
\ac{API} under the \texttt{segalubi} group, and the two prototypes
under the personal namespace of the author.

\section{Containerization: Docker and docker-compose}
\label{sec:docker}

\textbf{Docker}~\cite{docker} is used in two distinct roles in this
project. The first role is as a runtime for the \ac{CI} jobs: each
GitLab \ac{CI} job declared in this project runs inside a Docker
container based on the official \texttt{maven:3.9.6-eclipse-temurin-21}
image, which guarantees a clean and reproducible build environment
regardless of what is installed on the runner host. The second role
is as a local development aid: the two prototypes ship a
\texttt{docker-compose.yaml}~\cite{composeSpec} that stands up a
PostgreSQL instance, and, in the case of the first prototype, a
pair of Payara Server containers for the dual-node Arquillian tests
described in Section~\ref{sec:javaeeapp}.

For the production \ac{API}, Docker is intentionally \emph{not} used
to package the application: the laboratory's deployment model is to
copy the \ac{WAR} onto an existing Payara installation, and a
container image would not bring any benefit while adding an
additional moving part. This is briefly revisited as a possible
future evolution in Chapter~\ref{chap:conc-trab-futuro}.

\section{Network Access: The Laboratory \ac{VPN}}
\label{sec:vpn}

The production database of the laboratory is reachable on the
internal network address \texttt{192.168.168.30:5432} and is not
exposed to the public Internet. Two consequences follow. First, the
GitLab Runner that executes the differential tests against the real
database must itself be inside the laboratory's network, this is
why the runner is tagged \texttt{ubi-runner-.31} and pinned to a
specific machine. Second, developers working from outside the
laboratory have to connect to the network through a \ac{VPN}
configuration maintained by the laboratory staff (Luís Carvalho).

The technical choice of \ac{VPN} technology is outside the scope of
this project, the laboratory uses an existing solution, but
modern alternatives such as WireGuard~\cite{wireguard} were
considered briefly as a lighter-weight option for one-off remote
testing sessions.

\section{Conclusion}
\label{chap3:sec:concs}

The technology stack on which this project is built is in large part
imposed by its environment: Jakarta EE on Payara, PostgreSQL on the
internal network, GitLab as the self-managed \ac{CI} platform. Where
there was room for choice, between Jakarta EE and Spring Boot,
between Payara Server and Apache Tomcat, between PostgreSQL and
MySQL, between JUnit 5 and TestNG, the decisions documented in
this chapter were guided by a preference for fewer moving parts, by
the alignment with what the laboratory already runs in production
and by the desire to keep the differential testing harness as simple
as possible. The next chapter describes the iterative methodology by
which these technologies were combined into a working solution.
