\chapter{Conclusions and Future Work}
\label{chap:conc-trab-futuro}

\section{Main Conclusions}
\label{sec:conc-princ}

The work described in this report set out to address a very concrete
problem of the \ac{SEGAL} laboratory: how to refactor a large,
several-years-old Jakarta EE \ac{API} without silently breaking the
\ac{EPOS} clients that depend on it. The answer proposed and built
in the course of this project takes the form of a three-stage,
\ac{CI}/\ac{CD}-driven testing framework that for every commit
produced by the refactor compares the resulting build against the
immediately preceding one, both at the level of internal Java
classes and at the level of the \ac{HTTP} responses produced by the
\ac{API}.

Several conclusions emerge from the work.

\paragraph{Differential testing fits a refactor naturally.}
Asserting that a given input produces a given output is a brittle
contract during a refactor, because every legitimate cosmetic change
to a response field requires updating dozens of fixtures. Asserting
that the new build produces \emph{the same} output as the old one
sidesteps the problem entirely: a legitimate cosmetic change appears
once in the diff report, can be approved, and the next pipeline run
treats it as the new baseline. This is the property that allowed
the framework to be useful from the first commit it ran on, without
a long ramp-up period of fixture creation.

\paragraph{Two layers of comparison are better than one.}
The code-level harness (\texttt{DiffHarness.java}) and the
\ac{HTTP}-level harness (\texttt{HttpDiff.java}) catch different
classes of regressions. The first catches behavioural changes in
focused helper classes (\texttt{Validate}, \texttt{Shape}) before
they propagate into endpoint responses, with very fast feedback and
without the cost of booting an application server. The second
catches integration-level regressions that only manifest once a
real request hits a real database. Either one in isolation would
have left blind spots; together they cover the spectrum that the
laboratory cares about.

\paragraph{A self-hosted pipeline is not optional in this context.}
Because the production database lives on the internal network of
the laboratory and is not exposed to the public Internet, only a
runner that itself sits inside that network can exercise the
\ac{API} end-to-end against it. The decision to self-manage both
the GitLab control plane and the GitLab Runner host therefore
followed from the data-locality constraint, not from a preference
for self-hosting per se. The same architecture would not be
appropriate for an open-source project served by gitlab.com's
shared runners.

\paragraph{The two prototypes paid for themselves.}
In retrospect, having spent the first few weeks of the project on
\texttt{java-ee-app}, a deliberately trivial application whose
only purpose was to harden the pipeline, and then on
\texttt{syncspace-jakarta-app}, a deliberately fuller application
whose only purpose was to exercise Jakarta EE 11 features in
isolation, saved significant time later, when the production
\ac{API} had to be addressed. Each prototype answered a small,
well-bounded set of questions in an environment in which mistakes
were cheap. Without them, every one of those questions would have
had to be answered against the production code base, where mistakes
are not.

\paragraph{The framework reveals what a human review would miss.}
The clearest evidence of value, beyond any abstract argument, is the
list of commits in the repository in which a refactor was proposed,
the diff harness flagged a behavioural change, the change was
investigated, and the refactor was reverted. The most representative
example, discussed in Section~\ref{sec:reports}, is the named-constants
refactor of the latitude/longitude bounds: a textbook clean-up that
introduced a subtle off-by-one in the boundary check and that no
reasonable amount of code review would have caught at the rate the
laboratory commits. That alone is, in the author's view, the
justification for the framework.

\section{Future Work}
\label{sec:trab-futuro}

A number of items were deliberately left out of the scope of this
project, either because they were not strictly necessary to meet
the objectives, or because they would have expanded the project
beyond what could reasonably be addressed in the time available.

\paragraph{\ac{OpenAPI}-driven test case generation.}
The list of \ac{HTTP} cases exercised by \texttt{HttpDiff.java} is
currently maintained by hand in \texttt{http-diff-cases.json}. The
\ac{OpenAPI} 3.1.0 document shipped inside the \ac{WAR} already
describes every endpoint and the shape of its parameters; a future
evolution of the framework could derive the case list automatically
from the specification, ensuring that newly added endpoints are
exercised by the diff from the moment they are merged, with no need
to update the test fixture.

\paragraph{Schema diffing for the \ac{OpenAPI} document itself.}
A different but adjacent use of the same document is to diff the
\ac{OpenAPI} of the baseline against that of the \ac{HEAD}: any
breaking change to the contract (a removed endpoint, a changed
type, a tightened validation) would be reported as a separate kind
of regression. This would complement the behavioural diff with a
contract diff.

\paragraph{Containerization of the production \ac{API}.}
The production application is currently deployed by copying the
\ac{WAR} onto an existing Payara installation. Packaging the
application as a Docker image, with a Payara base image and a small
amount of entry-point configuration, would simplify the migration
to a containerised deployment platform if the laboratory ever
chooses to move in that direction, and would also remove the
ambiguity between ``the \ac{WAR} that was tested'' and ``the
\ac{WAR} that was deployed'' that occasionally arises with the
current model.

\paragraph{Performance regression testing and BRIN indexing.}
The current diff harnesses focus on behavioural equivalence and
ignore response time. A natural extension is to record the wall-clock
duration of each \ac{HTTP} case for both baselines and \ac{HEAD}
across both database fixtures (\texttt{DB\_3.0\_J} and
\texttt{DB\_3.0\_F}) and to flag cases in which the new version is
significantly slower than the old one. The threshold would
necessarily be statistical rather than absolute, because the runner
shares resources with other jobs; a small benchmarking subsystem
would be required. The first concrete consumer of such a subsystem
is already lined up: the planned introduction of BRIN indexes on the
date columns of \texttt{rinex\_file} and
\texttt{highrate\_rinex\_file} (Section~\ref{sec:epos-db}) needs
exactly this kind of empirical evidence to justify itself against
the larger \texttt{F} fixture.

\paragraph{PostGIS migration for spatial queries.}
Bounding-box and polygon checks are currently performed in Java by
the helpers of the \texttt{Geometry} package. Migrating those checks
to native PostGIS operators would let the database planner reason
about spatial selectivity and is the natural complement to the BRIN
work above. The expectation is that correctness would not change
significantly (the existing checks are already faithful to the
specification) and that the main benefit would be query plan
quality on the \texttt{DB\_3.0\_F} fixture, which makes this an
ideal target for the same differential framework: the same diff
cases should pass byte-for-byte against the PostGIS-enabled version,
and only the timing distribution should change.

\paragraph{Snapshot testing of representative responses.}
Not every endpoint is amenable to a moving baseline; some responses
should remain stable across many releases (for example, the
\texttt{/api/ping} response, or the \ac{OpenAPI} document itself).
A small set of snapshot tests, in addition to the differential ones,
would protect those invariants and would explicitly mark them as
``known stable'' rather than ``allowed to drift''.

\paragraph{Generalisation to other laboratory services.}
The framework has been designed against the specifics of
\texttt{api\_v4.0} but its architecture is not specific to that
\ac{API}: the dual-Payara-Micro approach, the worktree-based
baseline, the \ac{JSON}-driven case list and the \ac{JUnit}/\ac{CSV}
reporting could in principle be reused by any other Jakarta EE
service of the laboratory. Reorganising the harness code into a
small library, with the application-specific bits factored into
a clean configuration interface, would make this reuse
straightforward, and is a natural next step for a continuation of
the work.

\paragraph{Hardening of operational aspects.}
Finally, several operational aspects of the framework deserve more
work than was possible inside the scope of this project: failure
recovery in the harness (retries on transient network errors,
clean termination of orphan Payara Micro processes), more granular
\ac{VPN} routing, automated provisioning of the GitLab Runner host
through configuration management, and a richer dashboard on top of
the diff reports collected over time. None of these are blocking
for the day-to-day use of the framework, but each would make it
more comfortable to live with for the years to come.

\bigskip

Overall, the framework delivered by this project addresses the
problem that motivated it, has already proved its value on a real
regression caught during the refactor, and lays the groundwork for
the laboratory to continue evolving the \ac{API} with a much higher
degree of confidence than was previously possible. The items listed
in this section are natural next steps; none of them invalidates
the design that was put in place.
