\chapter{Introduction}
\label{chap:intro}

\section{Framing}
\label{sec:amb}

This document describes the work developed in the context of the
final-year project of the Computer Science and Engineering programme of
the \ac{UBI}, under the project code PC26-4 of the 2025/2026 academic
year. The project was proposed by, and carried out at, the \ac{SEGAL}
laboratory of \ac{UBI}, under the supervision of Professor Paul Andrew
Crocker and the co-supervision of Vasco Senra, with day-to-day
infrastructure support from Luís Carvalho and Fernando Geraldes
(responsible for the laboratory's virtual machines, internal network
and \ac{VPN} access).

\ac{SEGAL} is the research laboratory of \ac{UBI} that operates and
maintains a \ac{GNSS} data and metadata service contributing to the
\ac{EPOS}, a pan-European research infrastructure focused on solid Earth
science~\cite{epos,eposGnss}. As part of the \ac{EPOS} Thematic Core
Service for \ac{GNSS} data and products, \ac{SEGAL} runs a Jakarta EE
\ac{REST} \ac{API} that exposes \ac{GNSS} station metadata and
observation files (in the \ac{RINEX} format~\cite{rinex}) to scientists,
operators and downstream services across Europe.

The \ac{API} in question has been in operation, in various forms, for
several years. Over that period the code base accumulated significant
technical debt: methods of five thousand lines, files of more than
thirty thousand lines, duplicated logic copy-pasted across resources,
and a long tail of patches added by successive developers and
contributors to handle individual edge cases. A complete refactor of the
\ac{API} was therefore initiated, in which large parts of the
application were rewritten from scratch on top of Jakarta EE 11 and the
Payara Platform~\cite{jakartaee,payara}, while a small core of well
tested logic was carried over from the previous version.

A refactor of this magnitude poses a very specific risk: any seemingly
innocuous change in a query, a response field or a validation rule can
produce subtle, hard to detect regressions that only manifest when a
particular client, a particular station or a particular file is
requested. Until this project, validation of those changes was carried
out manually, a developer would deploy the new version, run a few
requests by hand, and compare the resulting payloads against the
previous one. This is precisely the situation against which this work
is positioned.

\section{Motivation}
\label{sec:mot}

The motivation for this project is the gap between the pace at which
the new \ac{API} is being developed and the rate at which a human can
reasonably validate that each new version behaves exactly like the
previous one on the cases that already worked. Manual testing of a
\ac{REST} \ac{API} that returns thousands of records, in several
formats (\ac{JSON}, \ac{XML}, \ac{CSV}) and through dozens of query
combinations, is both error prone and slow. Worse still, it does not
scale to the cadence of a continuous refactor where commits land
several times a day.

The need is twofold. On the one hand, the laboratory wanted an
\emph{automated} mechanism to compare every new build of the \ac{API}
against the previous one and flag behavioural divergences early, before
they reach the production deployment. On the other hand, the laboratory
also wanted to bring this mechanism inside a real \ac{CI}/\ac{CD}
pipeline running on infrastructure that the laboratory itself controls:
a self-managed GitLab instance, paired with one or more GitLab
Runners hosted inside the laboratory's network. The combination of the
two is what turns a one-off testing script into a sustainable practice.

In addition to the immediate value for the \ac{EPOS} \ac{API}, the
exercise of designing, prototyping and integrating such a framework was
also seen as a vehicle for the student to gain hands-on experience with
modern Jakarta EE features, with \ac{CI}/\ac{CD} pipelines, with
self-hosted infrastructure, and with the practice of regression testing
against a moving baseline, skills that are directly relevant beyond
the scope of this particular project.

\section{Objectives}
\label{sec:obj}

In line with the project proposal, the work described in this report
pursues the following objectives:

\begin{enumerate}
  \item To design and implement an \emph{automated regression testing
        framework} for a Jakarta EE \ac{REST} \ac{API} running on the
        Payara Platform, capable of comparing the behaviour of two
        builds of the application across multiple response formats
        (\ac{JSON}, \ac{XML}, \ac{CSV});
  \item To integrate that framework into a self-managed GitLab
        \ac{CI}/\ac{CD} pipeline, with dedicated GitLab Runners hosted
        inside the laboratory's network and with appropriate access to
        the laboratory's databases over a \ac{VPN};
  \item To validate the proposed approach on two prototype Jakarta EE
        applications, one focused on the \ac{CI}/\ac{CD} pipeline
        itself, the other focused on Jakarta EE~11 features, before
        applying it to the production \ac{API};
  \item To produce a structured comparison between consecutive versions
        of the production \ac{API}, both at the level of internal
        validation/geometry logic and at the level of \ac{HTTP}
        responses, with reports that are consumable both by a human
        developer and by the \ac{CI} system;
  \item To document the architecture, usage, limitations and operating
        procedure of the framework, in such a way that the laboratory
        can continue to use and evolve it after the end of this
        project.
\end{enumerate}

\section{Main Contributions}
\label{sec:contrib}

The main contributions of this work, expanded in detail in the
subsequent chapters, can be summarised as:

\begin{itemize}
  \item A reusable, three-stage GitLab \ac{CI}/\ac{CD} pipeline template
        for Jakarta EE applications deployed on Payara, with caching,
        artifact management and JUnit-compatible reporting;
  \item A code-level differential test harness
        (\texttt{DiffHarness.java}) that loads the current build and
        a baseline build of a Java class in two isolated class
        loaders and compares their outputs on a shared input grid;
        the harness is invoked twice by the pipeline, against two
        baselines (the immediate parent commit
        \texttt{HEAD\textasciitilde 1} and a fixed reference commit
        whose \ac{ID} is provided through the
        \texttt{DIFF\_FIXED\_BASELINE} \ac{CI}/\ac{CD} variable),
        producing a true three-way comparison per pipeline run;
  \item A complementary \ac{HTTP}-level differential test harness
        (\texttt{HttpDiff.java}) that boots one instance of Payara
        Micro per available \ac{WAR}, the \ac{HEAD} build and the
        two baselines, on different ports, replays a configurable
        suite of \ac{HTTP} cases against every instance and emits one
        report per baseline so that per-commit regressions and
        accumulated drift remain distinguishable;
  \item Two prototype Jakarta EE 11 applications, a minimal
        ``hello world'' service used to harden the pipeline itself,
        and a small room-booking \ac{API} used to evaluate Jakarta EE
        features in isolation, both reusable as didactic examples
        for future students of the laboratory.
\end{itemize}

\section{Organization of the Document}
\label{sec:organ}

In order to reflect the work carried out, this document is structured
into six chapters, each serving a specific purpose:

\begin{itemize}
  \item \textbf{Chapter 1 -- Introduction.} Frames the project in its
        institutional and technical context, presents the motivation
        and objectives, lists the main contributions and outlines the
        organization of the document.
  \item \textbf{Chapter 2 -- State of the Art.} Reviews the main
        families of test frameworks and \ac{CI}/\ac{CD} systems
        relevant to a Jakarta EE \ac{API}, and surveys existing
        approaches to differential testing of services, in order to
        situate the proposed solution.
  \item \textbf{Chapter 3 -- Technologies and Tools Used.} Describes
        the concrete stack of technologies adopted during the project:
        \ac{API} styles and specification formats, the Java/Jakarta
        ecosystem, \ac{SQL} databases, application servers, version
        control, containerization and networking.
  \item \textbf{Chapter 4 -- Methodology and Architecture.} Explains
        the iterative methodology, two prototype applications
        followed by the production application, and the overall
        architecture of the testing framework, with particular focus
        on the first prototype (\texttt{java-ee-app}) and the
        \ac{CI}/\ac{CD} patterns that were validated on it, as well
        as the second prototype (\texttt{syncspace-jakarta-app}) used
        for Jakarta EE 11 feature exploration.
  \item \textbf{Chapter 5 -- Use Case: EPOS API.} Describes the
        application of the methodology to the production \ac{API}
        (\texttt{api\_v4.0}): the \ac{EPOS} context, the structure of
        the application, the three-stage \ac{CI}/\ac{CD} pipeline, the
        two differential harnesses and the corresponding reports.
  \item \textbf{Chapter 6 -- Conclusions and Future Work.}
        Summarises the results, reflects on what was learned during
        the project and proposes directions for future work, in
        particular the items left out of scope.
\end{itemize}

\section{Repository Access}
\label{sec:repos}

The source code of all three applications discussed in this report is
hosted on GitLab. The production \ac{API}
(\texttt{api\_v4.0})~\footnote{\url{https://gitlab.com/segalubi/api_v4.0}}
is maintained inside the \texttt{segalubi} group and access is
restricted to laboratory members. The two prototype applications are
published in the author's personal namespace and can be consulted
publicly:

\begin{itemize}
  \item \texttt{java-ee-app}: \url{https://gitlab.com/PatalJunior/java-ee-app}
  \item \texttt{syncspace-jakarta-app}: \url{https://gitlab.com/PatalJunior/syncspace-jakarta-app}
\end{itemize}

Each repository contains a \texttt{README.md} with build instructions,
and a \texttt{.gitlab-ci.yml} file (where applicable) describing the
pipeline discussed in Chapters~\ref{chap:metodo-arq} and
\ref{chap:caso-epos}.
